---
title: "R Notebook"
output: html_notebook
---

Load the packages and the dataset itself

```{r}
library(tidyverse)
library(readr)
gun_deaths_cleaned <- read_csv("Downloads/gun_deaths_cleaned.csv")
View(gun_deaths_cleaned)
gun_deaths <- gun_deaths_cleaned
```

1.What is the total number of firearm deaths recorded in the dataset from 2012 to 2014?

```{r}
count(gun_deaths)
```

Total \# of deaths from gunfire through 2012-2014 is 98,015.

2.How does the annual trend in firearm deaths change between 2012 and 2014? Is there a significant increase or decrease?

```{r}
gun_deaths %>% 
  filter(year== '2012') %>% 
  count()
gun_deaths %>% 
  filter(year== '2013') %>% 
  count()
gun_deaths %>% 
  filter(year== '2014') %>% 
  count()
```

In 2012,32615 died to firearm deaths In 2013,32730 died to firearm deaths In 2014,32670 died to firearm deaths There was a small increase followed by the a small decrease the following year.

3.Which category (homicide, suicide, accidental) accounts for the most firearm deaths?

```{r}
gun_deaths %>% 
  filter(intent== 'Homicide') %>% 
  count()
gun_deaths %>% 
  filter(intent== 'Suicide') %>% 
  count()
gun_deaths %>% 
  filter(intent== 'Accidental') %>% 
  count()
```

Suicides accounted for the most firearm deaths with 62,291 deaths during 2012-2014

4.What is the percentage of firearm suicides compared to total firearm deaths in 2014?

```{r}
gun_deaths %>% 
  filter(year== '2014',intent== 'Suicide') %>% 
  count()
21039/32670
```

Around 64%

5.Which age group has the highest number of firearm-related suicides?

```{r}
clean <- gun_deaths %>% 
  filter(intent== 'Suicide') %>% 
  count(age) %>% 
  summary(age)
```

Highest number of suicides came in at the age of 54 with 1369 deaths

6.How does firearm homicide affect different racial/ethnic groups? Which group has the highest homicide rate?

```{r}
gun_deaths %>% 
  filter(intent=="Homicide") %>% 
  group_by(race) %>% 
  count()
```

7.What is the proportion of male versus female victims in firearm deaths

```{r}
gun_deaths %>% 
  group_by(sex) %>% 
  count()

```

Women account for 14% of firearm deaths,men are at 86%.

8.Which racial/ethnic group shows the highest increase in gun-related deaths from 2012 to 2014?

```{r}
gun_deaths %>% 
  group_by(year,race) %>% 
  count()
```
White indiviudlas had the highest growth in gunfire related deaths.

9.Did the overall number of firearm-related deaths increase, decrease, or remain stable from 2012 to 2014?
```{r}
gun_deaths %>% 
  group_by(year) %>% 
  count()
```
Overall number stayed stable.

10.Is there a seasonal pattern in firearm deaths (e.g., higher in summer months or holidays)?
```{r}
ggplot<-gun_deaths %>% 
  group_by(month) %>% 
  count()

ggplot(data = ggplot) + geom_line(mapping = aes(x = month,y = n)) + labs(title = "Gun Deaths over Time", x = "Month (1-12)",y = "Number of Deaths") + scale_x_continuous(breaks = 1:12,labels = month.abb) + labs(title = "Line Chart Example",x = "Month (1-12)",y = "Number of Deaths")
```
Summer has the highest increase in gundeaths overall,while winter had the biggest decrease

11.How do firearm suicide rates among older adults (45–64) compare year-over-year?
```{r}
gun_deaths %>% 
  filter(intent=='Suicide',age<=64 & age>=45,year=='2012') %>% 
  group_by(year,age) %>% 
  count()
gun_deaths %>% 
  filter(intent=='Suicide',age<=64 & age>=45,year=='2013') %>% 
  group_by(year,age) %>% 
  count()
gun_deaths %>% 
  filter(intent=='Suicide',age<=64 & age>=45,year=='2014') %>% 
  group_by(year,age) %>% 
  count()
```
suicide deaths tend to peak during the ages of 47-54 with a somewaht constant number yearly.

12.Is there a correlation between economic conditions and firearm deaths in this period?
```{r}
gun_deaths%>%
  group_by(education) %>% 
  count()
```
People with lower education are responsible for more gundeaths.
